---
title: "Collecting Text Data"
---

## Sources for Media Research

Media and communication researchers draw on several types of text sources:

| Source | Examples | Access |
|---|---|---|
| News archives | LexisNexis, Factiva, ProQuest | Institutional subscription |
| Social media | X/Twitter, Reddit, YouTube comments | APIs (rate-limited) |
| Government / institutional | Hansard, press releases, court records | Often open |
| Web pages | News sites, blogs | Scraping (check ToS) |
| Broadcast transcripts | TV news, podcasts | Varies |

## Working with Structured Exports

Many archives export to CSV, TSV, or JSON. The tidyverse handles these cleanly:

```{r}
#| eval: false
library(tidyverse)

# CSV export from a news database
articles <- read_csv("data/articles.csv")

# Typical columns: headline, date, source, body
glimpse(articles)
```

## The Guardian API

The Guardian provides a free API. Register at <https://open-platform.theguardian.com/>.

```{r}
#| eval: false
library(httr2)
library(jsonlite)

get_guardian <- function(query, from_date, api_key, n = 50) {
  req <- request("https://content.guardianapis.com/search") |>
    req_url_query(
      q            = query,
      `from-date`  = from_date,
      `page-size`  = n,
      `show-fields` = "bodyText",
      `api-key`    = api_key
    )

  resp <- req |> req_perform() |> resp_body_json()
  resp$response$results |>
    map_dfr(~ tibble(
      id      = .x$id,
      date    = .x$webPublicationDate,
      headline = .x$webTitle,
      body    = .x$fields$bodyText %||% NA_character_
    ))
}

# articles <- get_guardian("climate change", "2024-01-01", Sys.getenv("GUARDIAN_KEY"))
```

## Reddit via the pushshift / Reddit API

```{r}
#| eval: false
library(httr2)

get_subreddit_posts <- function(subreddit, limit = 25) {
  request(glue::glue("https://www.reddit.com/r/{subreddit}/new.json")) |>
    req_url_query(limit = limit) |>
    req_headers(`User-Agent` = "text-analysis-book/0.1") |>
    req_perform() |>
    resp_body_json() |>
    pluck("data", "children") |>
    map_dfr(~ .x$data[c("title", "selftext", "author", "created_utc", "score")] |>
              as_tibble())
}
```

## Ethical and Legal Considerations

Before collecting data, consider:

- **Terms of service**: most platforms prohibit scraping or restrict API use to non-commercial research.
- **Privacy**: even publicly posted content may carry reasonable privacy expectations.
- **IRB / ethics review**: check whether your institution requires review for computational social science projects.
- **Data retention**: store only what you need; anonymise where possible.

## Corpus Design

A corpus is not just a collection of documents — it is a theoretically motivated sample. Think carefully about:

- **Population**: what universe of texts are you trying to represent?
- **Time window**: does the phenomenon have clear temporal boundaries?
- **Source selection**: are some outlets or platforms systematically excluded?
- **Language**: are you equipped to handle code-switching or multilingual content?

Good corpus design is as important as good analysis.
