---
title:Tidyverse I: tidy data
---

## Introduction to Tidyverse

The goal of this chapter is to get you acquainted with the
[Tidyverse](https://www.tidyverse.org/). Tidyverse is a collection of
packages that have been designed around a singular and clearly defined
set of principles about what data should look like and how we should
work with it. It comes with a nice introduction in the [R for Data
Science](https://r4ds.hadley.nz/) book, for which the digital version is
available for free. This chapter deals with most of the material in
chapter 5 of that book.

In this chapter, we’ll focus on working with data using the
`tidyverse` package. This package includes the `dplyr` (data-pliers)
packages, which contains most of the tools we’re using below, but it
also contains functions for reading, analysing and visualising data that
will be explained later.

### The tidyverse package

Tidyverse is a package (or actually a collection of packages).
This means that it's not included in R, but you need to install it separately.  You've probably already done that, but if not,
install it with the *packages* pane in Rstudio or call `install.packages("tidyverse")`. This is only needed once on your computer, so you don't need to include this code in your scripts.

What you do need to do every time is *activate* the package using
the `library` function:


```{r, eval=F}
library(tidyverse)
```


::: {.callout-note title="Common errors"}
The two most common mistakes in R are forgetting to install a package and forgetting to activate it. In both cases, you will get an *error message*: a text displayed in red that explains what went wrong. Don't panic if you see an error: you will quickly learn to recognize the most common errors, and even if you don't, copy pasting the error into a search or chat engine (or an email to your instructor) often gives you a quick solution.

As an example, try the following two calls:

```{r, eval=F}
library(nastyverse)
format_log(100)
```

Both will give you an error message! The first will tell you `there's not package called 'nastyverse'` -- this means that either you made a typing error, or you forgot to install the package. If you think the name is correct, a quick `install.packages` generally fixes the problem. The second will say `could not find function "format_log"`. Again, unless you made a typo, this generally means you forgot to activate a package, in this case the `scales` package, which you can fix with `library(scales)`. But the most important thing is to not panic, read the error message, and either fix it or search for a solution.
:::

## Reading Data

As in most packages, the functionality in dplyr is offered through
functions. In general, a function can be seen as a command or
instruction to the computer to do something and (generally) return the
result.

In the tidyverse packages, almost all `functions` primarily operate on
data sets, for example for filtering and sorting data.

With a data set we mean a rectangular data frame consisting of rows
(often items or respondents) and columns (often measurements of or data
about these items). These data sets can be R `data.frames`, but
tidyverse has its own version of data frames called `tibble`, which is
mostly the same as a data frame but is more efficient and somewhat
easier to use.

### Creating a tibble

As a very simply example, the following code creates a tibble containing
respondents with a second column listing their gender:

``` r
respondents <- tibble(resp_id = c(1,2,3),
                      gender = c("M","M","F"))
```

Let’s say our respondents are 24, 32 and 27 years old. Can you add a
column called ‘age’ that specifies this age for the respondents?

``` r
library(tidyverse)
respondents <- tibble(resp_id = c(1,2,3),
                      gender = c("M","M","F"))
respondents
```

### A note on functions and assignment

In the exercise above, you used the `tibble` function to create a new
data frame, which you then *assigned* to the respondents object. If you
would not have assigned it to an object, R would simply print the result
of the function on the screen and then forget all about it. So, whenever
you want R to remember the result of a function, you need to assign it
to an object and give it a name. You can also re-use an existing name,
in which case the object will be over-written by the new result.

When writing R code, 90% of your lines will follow this pattern of
calling a function on some arguments (e.g. the data), and assigning it
to an object:

``` r
my_object <- some_function(arguments)
```

### Reading data with `read_csv`

The example above manually created a data set, but in most cases you
will start with data that you get from elsewhere, such as a csv file
(e.g. downloaded from an online dataset or exported from excel) or an
SPSS or Stata data file.

Tidyverse contains a function `read_csv` that allows you to read a csv
file directly into a data frame. You specify the location of the file,
either on your local drive or directly from the Internet!

The example below downloads an overview of gun polls from the [data
analytics site 538](https://fivethirtyeight.com/), and reads it into a
tibble using the read_csv function:

``` r
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
gunpolls <- read_csv(url)
gunpolls
```

(Note that if you run this in RStudio, you will probably see a (red)
message in the console. This is simply telling you how each column was
parsed (as a number, text, date, etc.), and is generally safe to ignore)

The tibble shows the first ten rows of the data set, and for each column
the data type is also mentioned: `<dbl>` stands for double, which is a
*numeric* value; `<chr>` stands for character or *textual* data.

Note that on your own computer, R will print a message if not all rows
or columns could be printed. If you want to browse through your data,
you can also click on the name of the data.frame (gunpolls) in the
top-right window “Environment” tab or call `View(gunpolls)`.

### Exercise: downloading data

A csv file of the political polls just after the fall of the Dutch
cabinet Rutte IV can be found at <https://i.amcat.nl/peiling.csv>. Can
you download that file and assign it to an object called `polls`?

``` r
library(tidyverse)
polls <- ____
polls
```

## Selecting and renaming columns

In the previous section, you managed to download data from the Internet.
Often, this data needs to be cleaned before we can use it for analysis.

R Tidyverse has a number of functions that can be used for data
cleaning. As a first step, let’s learn how to use `select` and `rename`
to change the *columns* of a data frame.

In this section, we will continue with the gun polls data we used above:

``` r
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
gunpolls <- read_csv(url)
gunpolls
```

### Selecting columns with `select`

This contains a number of columns that we might not be interested in.
Using `select`, we can select only the columns we are interested in:

``` r
select(gunpolls, Population, Support, Pollster)
```

You can also specify a range of columns, for example all columns from
Question to Support:

``` r
select(gunpolls, Question:Support)
```

Finally, it is also possible to drop columns by placing a minus sign in
front of the column name. For example, this will drop the URL column and
keep all other columns:

``` r
select(gunpolls, -URL)
```

### Renaming columns and non-standard names

In R, it is easiest if a column name does not contain any spaces or
other special characters. To select such columns, you need to use
‘backticks’ around the name (`` `like this` ``). Because this is
inconvenient, is is often better to rename such columns so they no
longer contain a space. Note that instead of a space, you can use an
underscore (`_`) if desired.

In R, you can use the `rename` function to change the name of one or
more variables:

``` r
rename(gunpolls, rep = `Republican Support`, dem = `Democratic Support`)
```

Very often, a first step in data cleaning is renaming some columns and
dropping others. This can be combined in a single step by renaming in
the `select` function:

``` r
select(gunpolls, Question, rep = `Republican Support`, dem = `Democratic Support`)
```

Note the difference between the two functions: `rename` keeps all
variables that are not mentioned, while `select` only keeps the
variables that you selected.

### Note: Assigning results to objects

In the examples above, the result of the select or rename was not
assigned to an object. This means that the results are printed on the
screen, but not remembered. In other words, the gunpolls object will
contain the same data after running the command.

To store the results of select or rename (or any other function), you
need to *assign* it to an object and give it a name. This can be a new
name, in which case a *copy* of the object will be made, so both objects
can be used later. After running the code below, both the full
`gunpolls` object (with all columns), and the `gunpolls_clean` object
will be remembered by R, so you can choose to continue analysing either.
This is most useful if you want to do some analyses on the subset of the
data, but also need to do other analyses on a different subset later.

``` r
gunpolls_clean <- select(gunpolls, Question, rep = `Republican Support`, dem = `Democratic Support`)
gunpolls_clean
```

If you assign it to the same name, it will overwrite the data:

``` r
gunpolls <- select(gunpolls, Question, rep = `Republican Support`, dem = `Democratic Support`)
gunpolls
```

This is useful if you will not need to data you discarded. To summarize:

``` r
some_function(data)          # <-- print on screen and forget
data2 <- some_function(data) # <-- create a new object and keep both
data <- some_function(data) # <-- overwrite the old object
```

### Exercise

Using the gun polls data from above, rename the `Republican Support` and
`Democratic Support` columns to `rep_support` and `dem_support`, and
select only the Question, Pollster, and republicat/democratic support
columns. Assign the result to the `gunpolls` object

``` r
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
gunpolls <- read_csv(url)
___

gunpolls  # prints the result on screen
```

## Subsetting (filtering) data

In the previous section we looked at selecting columns with the `select`
function. Another common task is selecting *rows*, which can be done
with the `filter` function.

This section also uses the gun polls data we used above:

``` r
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
gunpolls <- read_csv(url)
gunpolls
```

### The `filter` function

The `filter` function can be used to select a subset of rows. In the
guns data, the `Question` column specifies which question was asked. We
can select only those rows (polls) that asked whether the minimum
purchase age for guns should be raised to 21:

``` r
age21 <- filter(gunpolls, Question == 'age-21')
age21
```

This call is typical for a tidyverse function: the first argument is the
data to be used (`gunpolls`), and the remaining argument(s) contain
information on what should be done to the data (selecting purchase age
should be raised to 21).

Note the use of `==` for comparison: In R, `=` means assingment and `==`
means equals. Other comparisons are e.g. `>` (greather than), `<=` (less
than or equal) and `!=` (not equal). You can also combine multiple
conditions with logical (boolean) operators: `&` (and), `|` or, and `!`
(not), and you can use parentheses like in mathematics.

So, we can find all surveys where support for raising the gun age was at
least 80%:

``` r
filter(gunpolls, Question == 'age-21' & Support >= 80)
```

Note that this command did not assign the result to an object. So, as
explained above, the result is only displayed on the screen but not
remembered. This can be a great way to quickly inspect your data, but if
you want to continue analysing this subset you need to assign it to an
object, e.g. using `subset <- filter(...)`

### Removing missing values

Often, data contains *missing values*, for example survey questions that
were not answered or other data that is missing or unknown. In R, such
values are called `NA` (Not Available), and any comparison with a
missing value are automatically missing as well. To check if a value is
missing, you should use the function `is.na(column)`.

For example, let’s look at a version of the gun polls data which has
some missing values:

``` r
library(tidyverse)
gunpolls_dirty <- read_csv("https://cssbook.net/d/guns-polls-dirty.csv")
head(gunpolls_dirty)
```

As you can see, the Support column has a missing value for CBS news. To
remove rows where Support is missing, `filter` the dataset so only
non-missing rows are kept, i.e. where `!is.na(Support)` (you can read
`!is.na` as ‘is not missing’):

``` r
filter(gunpolls_dirty, !is.na(Support))
```

As a shortcut to remove all rows where any column is missing, you can
use the `na.omit` function:

``` r
na.omit(gunpolls_dirty)
```

### Exercise: Filtering data

Let’s go back to the original gunpolls data. Can you select the polls
about stricter gun laws (`stricter-gun-laws`) where the overall suppport
was at least 70%, and assign that selection to a new object called
`strict_polls`?

``` r
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
gunpolls <- read_csv(url)
___
strict_polls
```

## Changing or computing values

As a final part of this tutorial, let’s look at how you can change
values using the `mutate` function:

### The `mutate()` function

The mutate function makes it easy to create new variables or to modify
existing ones. For those more familiar with SPSS, this is what you would
do with `compute` and `recode`. As a simple example, let’s start again
with the 538 gun polls data used above:

``` r
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
gunpolls <- read_csv(url)
gunpolls <- select(gunpolls, Question, Support,
                   rep=`Republican Support`, dem=`Democratic Support`)
```

For example, to transform the Support percentage (0-100) to a proportion
(0-1), we can divide the column by 100 using `mutate`:

``` r
mutate(gunpolls, Support_Prop = Support / 100)
```

The syntax of mutate is similar to that of `filter()` and `select()`:
The first argument is the data frame to mutate, and then any number of
additional arguments can be given to perform mutations. The mutations
themselves are named arguments, in which you can provide any
calculations using existing columns such as `Support`. The result of the
calculatoin will then be stored in the name of the argument, in this
case `Support_Prop`. You can also overwrite columns by storing the
result with an existing column’s name.

For calculations, you can use mathematical symbols such as `+`, `/`
(division), and `*` (multiplication). There are also many useful
functions that you can use in calculations, such as `round()` to round
values, `as.numeric()` to change a text column to a numberic column, or
`abs()` to take the absolute (positive) value of numbers. Feel free to
play around a bit in the code above.

### Exercise

Using the `gunpolls` data as defined above, can you create a new column
`party_diff` that contains the absolute value of the difference between
republican and democratic support? That is, if republican support is 61
and democratic support is 86 (or the other way around), the difference
should be 25. Overwrite the gunpolls object with the result (i.e. add
the column to the existing data set).

``` r
gunpolls <- mutate(gunpolls, ____)
gunpolls
```
