---
title: "Topic Modeling"
---

## Latent Dirichlet Allocation

Latent Dirichlet Allocation (LDA) is a generative probabilistic model that represents documents as mixtures of topics, and topics as distributions over words [@blei2003latent]. It is unsupervised: you do not provide labels.

**Intuition**: if a corpus contains articles about politics, economics, and sports, LDA should recover three clusters of co-occurring words, even without being told the categories.

## Fitting an LDA Model

```{r}
#| echo: true
#| eval: false
library(tidytext)
library(topicmodels)
library(dplyr)
library(ggplot2)

# Assume `articles` is a data frame with columns: doc_id, text
dtm <- articles |>
  unnest_tokens(word, text) |>
  anti_join(stop_words) |>
  count(doc_id, word) |>
  cast_dtm(doc_id, word, n)

lda_model <- LDA(dtm, k = 6, control = list(seed = 42))
```

```{webr}
# Demonstration with a small simulated corpus
library(tidytext)
library(topicmodels)
library(dplyr)
library(ggplot2)

sim_texts <- tibble(
  doc = 1:12,
  text = c(
    "election vote party candidate ballot democracy poll",
    "minister parliament legislation bill policy vote government",
    "economy inflation gdp growth unemployment interest rate",
    "market stock trade investment finance bank growth",
    "climate temperature emissions carbon fossil fuel energy",
    "flood drought weather extreme temperature disaster",
    "film award actor director cinema screen performance",
    "music album concert tour artist singer record",
    "election candidate party poll vote campaign ballot",
    "parliament minister legislation government policy reform",
    "climate energy carbon emissions renewable solar wind",
    "economy bank market inflation interest gdp growth"
  )
)

dtm <- sim_texts |>
  unnest_tokens(word, text) |>
  count(doc, word) |>
  cast_dtm(doc, word, n)

lda_out <- LDA(dtm, k = 3, control = list(seed = 1))

# Top terms per topic
tidy(lda_out, matrix = "beta") |>
  group_by(topic) |>
  slice_max(beta, n = 6) |>
  ggplot(aes(beta, reorder(term, beta), fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ paste("Topic", topic), scales = "free_y") +
  labs(x = "β (word probability)", y = NULL,
       title = "Top words per LDA topic") +
  theme_minimal()
```

## Document-Topic Distributions

```{webr}
library(tidytext)
library(topicmodels)
library(dplyr)
library(ggplot2)

sim_texts <- tibble(
  doc = 1:12,
  text = c(
    "election vote party candidate ballot democracy poll",
    "minister parliament legislation bill policy vote government",
    "economy inflation gdp growth unemployment interest rate",
    "market stock trade investment finance bank growth",
    "climate temperature emissions carbon fossil fuel energy",
    "flood drought weather extreme temperature disaster",
    "film award actor director cinema screen performance",
    "music album concert tour artist singer record",
    "election candidate party poll vote campaign ballot",
    "parliament minister legislation government policy reform",
    "climate energy carbon emissions renewable solar wind",
    "economy bank market inflation interest gdp growth"
  )
)

dtm <- sim_texts |>
  unnest_tokens(word, text) |>
  count(doc, word) |>
  cast_dtm(doc, word, n)

lda_out <- LDA(dtm, k = 3, control = list(seed = 1))

tidy(lda_out, matrix = "gamma") |>
  mutate(document = as.integer(document)) |>
  ggplot(aes(factor(topic), gamma, fill = factor(topic))) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  labs(x = "Topic", y = "γ (document-topic proportion)",
       title = "Document-topic distribution") +
  theme_minimal()
```

## Choosing K

Selecting the number of topics is a modelling decision, not a statistical optimum. Common approaches:

- **Perplexity** on held-out data (lower = better fit, but may favour large *k*)
- **Coherence** (higher = more interpretable topics)
- **Substantive judgement**: do topics make sense to domain experts?

```{r}
#| eval: false
library(topicmodels)

perplexities <- sapply(c(2, 4, 6, 8, 10, 15, 20), function(k) {
  m <- LDA(dtm_train, k = k, control = list(seed = 42))
  perplexity(m, newdata = dtm_test)
})
```

## Structural Topic Models

The `stm` package extends LDA to allow document metadata (e.g., publication date, outlet type) to influence topic prevalence or content:

```{r}
#| eval: false
library(stm)

# out <- prepDocuments(documents, vocab, meta)
# model <- stm(out$documents, out$vocab, K = 10,
#              prevalence = ~ outlet + s(year),
#              data = out$meta)
# plot(model, type = "summary")
```

STM is particularly useful for testing whether topic prevalence differs by outlet or over time.

## References

@blei2003latent introduced LDA. @roberts2014structural describe STM. @grimmer2022text ch. 13 gives a social-science-oriented treatment.
