---
title: "Computational Content Analysis"
---

## From Manual to Automated Coding

Traditional content analysis uses human coders to apply a codebook to a sample of texts. Computational approaches extend this in two ways:

1. **Dictionary methods**: define categories by word lists, apply mechanically
2. **Supervised machine learning**: train a classifier on human-coded examples, apply to the full corpus

Both require a validated codebook and human-coded training/validation data.

## Dictionary-Based Classification

### Policy Issue Coding

The Comparative Manifesto Project and Policy Agendas Project provide validated category schemes. Here we build a simple custom dictionary:

```{webr}
library(quanteda)
library(quanteda.textstats)

media_dict <- dictionary(list(
  economic    = c("economy", "inflation", "unemployment", "gdp", "growth",
                  "market", "trade", "finance", "fiscal", "budget"),
  political   = c("parliament", "election", "party", "minister", "vote",
                  "government", "opposition", "policy", "legislation"),
  environment = c("climate", "emissions", "carbon", "pollution", "renewable",
                  "energy", "fossil", "temperature", "flood", "drought")
))

headlines <- c(
  h1 = "Government announces new fiscal policy amid rising inflation",
  h2 = "Climate summit: nations pledge to cut carbon emissions by 2030",
  h3 = "Opposition party wins election on economic reform platform",
  h4 = "Renewable energy investment reaches record high as fossil fuels decline",
  h5 = "Parliament debates new legislation on financial market regulation"
)

corp <- corpus(headlines)
toks <- tokens(corp, remove_punct = TRUE) |> tokens_tolower()

dfm(toks) |>
  dfm_lookup(media_dict) |>
  convert(to = "data.frame")
```

### Framing Analysis

```{webr}
library(quanteda)

framing_dict <- dictionary(list(
  conflict    = c("clash", "battle", "fight", "opponent", "attack",
                  "versus", "confrontation", "dispute", "war"),
  human_interest = c("family", "child", "community", "personal", "story",
                     "victim", "struggle", "hope", "life"),
  economic_consequence = c("cost", "billion", "million", "impact", "growth",
                           "loss", "profit", "economy", "financial")
))

articles <- c(
  a1 = "The clash between government and unions escalated as both sides refused to budge.",
  a2 = "Families struggle with rising costs as the economic impact of the crisis deepens.",
  a3 = "The dispute over billions in public funding pits the opposition against the minister.",
  a4 = "Personal stories of hope emerge from communities affected by the disaster."
)

corpus(articles) |>
  tokens(remove_punct = TRUE) |>
  tokens_tolower() |>
  dfm() |>
  dfm_lookup(framing_dict) |>
  convert(to = "data.frame")
```

## Supervised Classification

When dictionaries are insufficient, train a classifier on hand-coded examples.

```{r}
#| eval: false
library(tidymodels)
library(textrecipes)

# Assumes `labelled` has columns: text, label
set.seed(42)
split  <- initial_split(labelled, prop = 0.8, strata = label)
train  <- training(split)
test   <- testing(split)

rec <- recipe(label ~ text, data = train) |>
  step_tokenize(text) |>
  step_stopwords(text) |>
  step_tokenfilter(text, max_tokens = 1000) |>
  step_tfidf(text)

spec <- logistic_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet")

wf <- workflow() |>
  add_recipe(rec) |>
  add_model(spec)

folds <- vfold_cv(train, v = 5, strata = label)
tune_res <- tune_grid(wf, resamples = folds, grid = 20)

best  <- select_best(tune_res, metric = "roc_auc")
final <- finalize_workflow(wf, best) |> last_fit(split)
collect_metrics(final)
```

## Intercoder Reliability

Before using human labels as training data, establish that your codebook is reliable:

```{r}
#| eval: false
library(irr)

# Two coders applied scheme to 100 documents
# kappa >= 0.7 is a common threshold for acceptable reliability
ratings <- data.frame(
  coder1 = c(1,1,2,3,1,2,3,1,2,1),  # example
  coder2 = c(1,1,2,2,1,2,3,1,2,1)
)

kappa2(ratings)       # Cohen's κ for two coders
kripp.alpha(t(as.matrix(ratings)), method = "nominal")  # Krippendorff's α
```

## Validation Strategy

A robust validation protocol:

1. Set aside a **held-out test set** (never used during development)
2. Report precision, recall, and F1 per category
3. Manually inspect misclassifications — they reveal systematic errors
4. Check performance across subgroups (time periods, outlet types)
