---
title: "Word Embeddings and Semantic Analysis"
---

## Beyond Bag-of-Words

Bag-of-words models treat words as independent tokens. Word embeddings represent words as dense vectors in a continuous space, capturing semantic relationships: words used in similar contexts have similar vectors.

Key properties:

- *king* − *man* + *woman* ≈ *queen*
- *BBC* and *CNN* will be near each other
- Distances reflect semantic similarity

## Word2Vec and GloVe

The two dominant pre-trained embedding families:

| Model | Trained on | Dimensions | Use case |
|---|---|---|---|
| word2vec (Google) | Google News | 300 | General English |
| GloVe (Stanford) | Wikipedia + Gigaword | 50–300 | General English |
| fastText | Common Crawl | 300 | Subword-aware |

For domain-specific corpora (e.g., political news), training embeddings on your own data often outperforms off-the-shelf models.

## Using Pre-trained Embeddings

```{r}
#| eval: false
library(word2vec)

# Download GloVe vectors (large file, do once)
# https://nlp.stanford.edu/projects/glove/

model <- read.word2vec("glove.6B.100d.bin")

# Most similar words
predict(model, "journalist", type = "nearest", top_n = 10)

# Analogies
# king - man + woman
v <- word2vec_similarity(
  predict(model, c("king", "man", "woman"), type = "embedding"),
  predict(model, rownames(model$words), type = "embedding")
)
```

## Training Embeddings on Your Corpus

```{r}
#| eval: false
library(word2vec)

model <- word2vec(
  x         = your_corpus_sentences,  # character vector, one sentence per element
  type      = "cbow",
  dim       = 100,
  window    = 5,
  min_count = 5,
  iter      = 10,
  threads   = 4
)

write.word2vec(model, "models/corpus_w2v.bin")
```

## Measuring Semantic Change Over Time

A powerful media research application: track how the meaning of a term shifts across time periods.

```{r}
#| eval: false
# Train separate models per decade/period
# Align vector spaces using Procrustes alignment
# Compare nearest neighbours or cosine distances

# Conceptual workflow:
periods <- list(
  "1990s" = corpus_90s,
  "2000s" = corpus_00s,
  "2010s" = corpus_10s
)

models <- lapply(periods, \(corp) {
  word2vec(corp, dim = 100, iter = 10, min_count = 10)
})

# Then align and compare vectors for target terms
```

## Concept Mover's Distance

CMD measures how far documents are from a concept defined by seed words, using Earth Mover's Distance over word embeddings [@stoltz2019concept]:

```{r}
#| eval: false
library(text2map)

# Define concept pole
populism_seeds <- c("people", "elite", "corrupt", "establishment", "ordinary")

cmd_scores <- CMDist(
  dtm    = your_dtm,
  cw     = populism_seeds,
  wv     = glove_matrix
)
```

## Large Language Model Embeddings

Modern transformer-based embeddings (e.g., from OpenAI, sentence-transformers) capture contextual meaning more accurately than static word2vec/GloVe. They are computationally heavier but accessible via API:

```{r}
#| eval: false
library(httr2)

get_embedding <- function(text, api_key) {
  request("https://api.openai.com/v1/embeddings") |>
    req_auth_bearer_token(api_key) |>
    req_body_json(list(model = "text-embedding-3-small", input = text)) |>
    req_perform() |>
    resp_body_json() |>
    pluck("data", 1, "embedding") |>
    unlist()
}
```

## Further Reading

- @rodriguez2022word review word embeddings for social scientists.
- @kozlowski2019geometry demonstrate embedding-based cultural analysis.
- @stoltz2019concept introduce Concept Mover's Distance.
